---
title: "Aggregate Hydrofabric"
params:
  hf_path: "..."
  aggregate_outlets_file: "..."
  aggregate_distribution_file: "..."
  ideal_size_sqkm: ...
  min_length_km: ...
  min_area_sqkm: ...
---

### Run Aggregation Workflow

```{r}
library(hfutils)
library(hfaggregate)
```

There are two ways to aggregate the hydrofabric, the first is using user-defined points of interest (POIs). Since I don't have any POIs I'm just grabbing the hydrolocations within the reference hydrofabric. From what I can tell, this won't make any changes to the hydrofabric since no new POIs are included. It seems like this is a place the CIROH researcher may want to insert their own POIs that don't exist in the reference hydrofabric.

```{r}
pois  <- hfutils::as_ogr(params$hf_path, "hydrolocations") |> 
  dplyr::select(flowpath_id, poi_id, geom) |> 
  hfutils::st_as_sf()


outlet_net <- aggregate_to_outlets(gpkg = params$hf_path,
                                   pois = pois,
                                   outfile = params$aggregate_outlets_file
                                   )
```

The second way to aggregate the hydrofabric is by modifying attributes such as `ideal_size_sqkm`.

```{r}
# Build a uniformly sized distribution
aggregated <- aggregate_to_distribution(
  gpkg = params$hf_path,
  ideal_size_sqkm = params$ideal_size_sqkm,
  min_length_km = params$min_length_km,
  min_area_sqkm = params$min_area_sqkm,
  outfile = params$aggregate_distribution_file
)
```