---
title: "Reconcile HydroFabric Attributes"
date: "2026-05-04"
params:
  ngen_gpkg: "..."
  vpu_id: "..."
  out_nextgen: "..."
  reference_gpkg: "..."
  nwm_lakes_gpkg: "..."
---

```{r}
library(sf)
library(dplyr)
library(tidyr)
```

## Prepare the Output NextGen Geopackage

```{r}
# Copy the input geopackage to a new file, then add the attributes
file.copy(params$ngen_gpkg, params$out_nextgen)


```

## Fix The Divide Attributes

```{r}

# get the flowpath_id to toid mapping from the flowpaths layer
fp_toid <- st_read(params$ngen_gpkg, layer = "flowpaths") |>
  st_drop_geometry() |>
  select(flowpath_id, toid = flowpath_toid)

# compute divides with the correct id, divide_id, toid, and other attributes
# this does the following transformations:
#   - renames fp-43378 to id = wb-43378 (just swap prefix fp- → wb-)
#   - adds toid from the flowpaths layer using the flowpath_id
divides <- st_read(params$ngen_gpkg, layer = "divides") |>
  left_join(fp_toid, by = "flowpath_id") |>
  mutate(
    id    = gsub("^fp-", "wb-", flowpath_id),
    # ds_id is all-NA for most VPUs, which makes sf infer a logical/BOOLEAN
    # column instead of the expected REAL; force it explicitly so the
    # written type always matches the target schema regardless of content
    ds_id = as.double(ds_id)
  ) |>
  select(-flowpath_id)
#  select(id, divide_id, toid, ds_id, areasqkm, vpuid, type, has_flowline)

# add missing vpuid column
divides$vpuid <- params$vpu_id


```

## Fix the Flowpath Attributes

```{r}

# read the flowpaths layer and do the following transformations:
#  - rename fp-43378 to id = wb-43378 (swaps prefix fp to wb)
#  - add toid from the flowpaths layer using the flowpath_id
#  - drop columns that we no longer need.
#  - join stream order from the reference hydrofabric via member_comid, taking
#    the maximum streamorde across all member COMIDs for each aggregated flowpath

# build stream order lookup from reference hydrofabric: id -> streamorde
ref_streamorde <- st_read(params$reference_gpkg, layer = "flowpaths", quiet = TRUE) |>
  st_drop_geometry() |>
  select(id, streamorde) |>
  mutate(id = as.character(as.integer(id)))

# for each flowpath, split member_comid (comma-separated), join to ref stream order,
# and take the max across all member COMIDs
order_lookup <- st_read(params$ngen_gpkg, layer = "flowpaths", quiet = TRUE) |>
  st_drop_geometry() |>
  select(flowpath_id, member_comid) |>
  separate_rows(member_comid, sep = ",") |>
  mutate(member_comid = trimws(member_comid)) |>
  left_join(ref_streamorde, by = c("member_comid" = "id")) |>
  group_by(flowpath_id) |>
  summarize(order = max(streamorde, na.rm = TRUE), .groups = "drop") |>
  mutate(order = if_else(is.infinite(order), NA_real_, order))

flowpaths <- st_read(params$ngen_gpkg, layer = "flowpaths") |>
  mutate(
         id = gsub("^fp-", "wb-", flowpath_id),
         toid = flowpath_toid,
         # mainstem and poi_id come straight from the upstream build and are
         # sometimes entirely NA for a given VPU, which makes sf infer
         # BOOLEAN/logical columns instead of the expected REAL/TEXT types;
         # force them explicitly so the written type always matches the
         # target schema regardless of content
         mainstem = as.double(mainstem),
         poi_id   = as.character(poi_id),
         ) |>
  left_join(order_lookup, by = "flowpath_id") |>
  select(-flowpath_id, -flowpath_toid, -flowline_id, -member_comid)


# add missing vpuid column
flowpaths$vpuid <- params$vpu_id

```

## Fix the Nexus Attributes

```{r}

# read the nexus layer and do the following transformations:
#  - set id to be the same as the nexus_id column
#  - rename nexus_toid to toid, and swap the prefix from fp- to wb- (e.g. fp-43378 becomes wb-43378)
#  - drop columns that we no longer need
#  - drop duplicate id rows: the upstream build occasionally emits more than
#    one row for the same nexus_id (both truly identical duplicates, and in
#    some cases the same id reused at slightly different coordinates, e.g.
#    a shared "nex-0" terminal/no-downstream sentinel reused across many
#    unrelated outlets). Downstream routing tools (e.g. t-route) index their
#    network by this id and crash ("cannot reindex on an axis with duplicate
#    labels") if it isn't unique, so keep only the first occurrence of each id.
nexus <- st_read(params$ngen_gpkg, layer = "nexus") |> 
  mutate(
         id   = nexus_id,
         toid = gsub("^fp-", "wb-", nexus_toid)
  ) |>
  select(-nexus_id, -nexus_toid) |>
  distinct(id, .keep_all = TRUE)

# Add missing POI attributes which are defined in the network layer.
# rename the nexus_id column to match the flowpath_id column in the network layer,
# then join to get the poi_id
poi_lookup <- st_read(params$ngen_gpkg, layer = "network") |>
     dplyr::select(flowpath_id, poi_id) |>
     dplyr::distinct() |>
     dplyr::filter(!is.na(poi_id)) |>
     dplyr::mutate(
       nexus_id = gsub("^fp-", "nex-", flowpath_id),
       # poi_id comes through as TEXT from the upstream build; the target
       # schema expects an integer (MEDIUMINT) id, matching the NWM
       # lake/gage identifiers it's ultimately joined against
       poi_id   = as.integer(poi_id)
     ) |>
     dplyr::select(nexus_id, poi_id)
nexus <- left_join(nexus, poi_lookup, by = c("id" = "nexus_id"))

# add placeholder values for missing coumns
nexus$type   <- "nexus"
nexus$vpuid <- params$vpu_id

```

## Write the updated divides and flowpaths to the output NextGen Geopackage

```{r}

# build lookups from the reference hydrofabric network layer for:
#   mainstem/hf_mainstem (mainstemlp), topo
# the reference network has one row per flowline; aggregate to one per ngen flowpath
# via reference_id, taking max(mainstemlp) and mode(topo)
ref_net <- st_read(params$reference_gpkg, layer = "network", quiet = TRUE) |>
  st_drop_geometry() |>
  select(id, mainstemlp, topo) |>
  mutate(id = as.character(as.integer(id)))

ref_net_lookup <- st_read(params$ngen_gpkg, layer = "network", quiet = TRUE) |>
  st_drop_geometry() |>
  select(flowpath_id, reference_id) |>
  mutate(reference_id = as.character(as.integer(reference_id))) |>
  left_join(ref_net, by = c("reference_id" = "id")) |>
  group_by(flowpath_id) |>
  summarize(
    mainstem = max(mainstemlp, na.rm = TRUE),
    topo     = names(sort(table(topo), decreasing = TRUE))[1],
    .groups  = "drop"
  ) |>
  mutate(mainstem = if_else(is.infinite(mainstem), NA_real_, mainstem))

# build lookup for ds_id and type from the ngen divides layer
divide_lookup <- st_read(params$ngen_gpkg, layer = "divides", quiet = TRUE) |>
  st_drop_geometry() |>
  select(flowpath_id, ds_id, type)

# Fix the network layer:
#  - rename columns to match the hydrofabric data model
#  - swap fp- prefix to wb- on flowpath_id and flowpath_toid
#  - join ds_id, type, mainstem, hf_mainstem, topo
#  - add vpuid
#  - drop columns that are not needed
network <- st_read(params$ngen_gpkg, layer = "network") |>
  mutate(
    id   = gsub("^fp-", "wb-", flowpath_id),
    toid = gsub("^fp-", "wb-", flowpath_toid)
  ) |>
  rename(
    lengthkm     = flowpath_lengthkm,
    hf_id        = flowline_id,
    hf_lengthkm  = flowline_lengthkm,
    hf_hydroseq  = flowline_hydroseq,
    hf_source    = hl_source
  ) |>
  left_join(ref_net_lookup, by = "flowpath_id") |>
  left_join(divide_lookup,  by = "flowpath_id") |>
  mutate(
    hf_mainstem = mainstem,
    vpuid       = params$vpu_id,
    # poi_id comes through as TEXT from the upstream build; the target
    # schema expects an integer (MEDIUMINT) id
    poi_id      = as.integer(poi_id)
  ) |>
  select(-flowpath_id, -flowpath_toid, -incremental_area_id, -incremental_areasqkm,
         -hl_reference, -hl_class, -flowline_toid, -reference_id)

# the network layer should have one row per node in the flow network, not just
# one row per flowpath. Append a row per nexus (id = nexus id, toid = the
# downstream flowpath it flows into) so that tools which build the network
# topology graph directly from this layer (e.g. ngiab_data_cli) see the
# nexus -> flowpath edges, not just the flowpath -> nexus edges.
nexus_network_rows <- nexus |>
  st_drop_geometry() |>
  transmute(
    id                    = id,
    toid                  = toid,
    divide_id             = NA,
    ds_id                 = NA,
    mainstem              = NA,
    hydroseq              = NA,
    hf_source             = "NOAA Reference Fabric",
    hf_id                 = NA,
    lengthkm              = NA,
    areasqkm              = NA,
    tot_drainage_areasqkm = NA,
    type                  = "nexus",
    vpuid                 = vpuid,
    hf_hydroseq           = NA,
    hf_lengthkm           = NA,
    hf_mainstem           = NA,
    topo                  = "fl-nex",
    poi_id                = poi_id,
    hl_uri                = NA
  )

network <- bind_rows(network, nexus_network_rows)

# combining the flowpath rows (read from source, e.g. integer hydroseq) with
# the placeholder nexus rows above (bare `NA`, which is logical) can shift a
# column's inferred type during bind_rows (e.g. integer -> double). Enforce
# the target schema's types explicitly on the final combined table so this
# is correct regardless of how bind_rows() resolves each column.
network <- network |>
  mutate(
    ds_id       = as.double(ds_id),
    hydroseq    = as.integer(hydroseq),
    poi_id      = as.integer(poi_id)
  )

```

## Build the POIs Layer

```{r}

# build a lookup from poi_id to id (wb-), nex_id (nex-), and hf_id (flowline_id)
# using the raw (pre-prefix-swap) network layer
poi_id_lookup <- st_read(params$ngen_gpkg, layer = "network", quiet = TRUE) |>
  st_drop_geometry() |>
  select(flowpath_id, poi_id, hf_id = flowline_id) |>
  filter(!is.na(poi_id)) |>
  distinct(flowpath_id, poi_id, .keep_all = TRUE) |>
  mutate(
    id     = gsub("^fp-", "wb-",  flowpath_id),
    nex_id = gsub("^fp-", "nex-", flowpath_id),
    # poi_id comes through as TEXT from the upstream build; the target
    # schema expects an integer (MEDIUMINT) id, matching the NWM
    # lake/gage identifiers it's ultimately joined against
    poi_id = as.integer(poi_id)
  ) |>
  select(poi_id, id, nex_id, hf_id)

# build the pois layer from the (raw, pre-prefix-swap) network layer:
#  - poi_id: the POI identifier
#  - id: the flowpath id, wb- prefixed
#  - nex_id: the nexus id, nex- prefixed
#  - vpuid: the vpu id
pois <- poi_id_lookup |>
  mutate(vpuid = params$vpu_id) |>
  select(poi_id, id, nex_id, vpuid)

```

## Fix the Hydrolocations Layer

```{r}

# read the hydrolocations layer and do the following transformations:
#  - join id, nex_id, hf_id from the poi_id_lookup built above
#  - derive hl_link by stripping the {hl_type}- prefix from hl_reference
#  - derive hl_x, hl_y from the point geometry
#  - add vpuid
#  - drop hl_class, which is not present in the target schema
hydrolocations <- st_read(params$ngen_gpkg, layer = "hydrolocations") |>
  mutate(
    # poi_id comes through as TEXT from the upstream build; cast to integer
    # before joining so it matches poi_id_lookup's (also-integer) poi_id key
    poi_id = as.integer(poi_id)
  ) |>
  left_join(poi_id_lookup, by = "poi_id") |>
  mutate(
    hl_link = sub("^[^-]+-", "", hl_reference),
    hl_x    = st_coordinates(geom)[, 1],
    hl_y    = st_coordinates(geom)[, 2],
    vpuid   = params$vpu_id
  ) |>
  select(-hl_class)

```

## Build the Lakes Layer

```{r}

# read the NWM lakes geopackage, reproject to match the ngen geopackage CRS,
# and derive hf_id (from comid) and lake_x/lake_y from the transformed geometry.
# filter to the POIs present in this VPU using an inner join on poi_id.
lake_params <- st_read(params$nwm_lakes_gpkg, layer = "lakes", quiet = TRUE) |>
  st_transform(st_crs(hydrolocations)) |>
  rename(hf_id = comid) |>
  mutate(
    lake_x = st_coordinates(geom)[, 1],
    lake_y = st_coordinates(geom)[, 2]
  ) |>
  select(-lat, -lon, -time, -ascendingIndex, -crs, -x, -y)

# filter to POIs present in this VPU's pois layer, add vpuid
lakes <- lake_params |>
  # poi_id is already an integer (MEDIUMINT) in the NWM lakes source, matching
  # the target schema; cast explicitly so the join is robust even if pois$poi_id
  # or lake_params$poi_id's exact numeric subtype ever drifts
  mutate(poi_id = as.integer(poi_id)) |>
  inner_join(select(st_drop_geometry(pois), poi_id), by = "poi_id") |>
  mutate(vpuid = params$vpu_id)

```

## Aggregate the Flowpath (Channel) Attribute Tables to One Row per Flowpath

```{r}

# The upstream build's flowpath-attributes / flowpath-attributes-ml tables are
# still at raw NHDPlus-flowline granularity: many rows can share the same
# ngen flowpath `id` (fragments of the original NHD reach network that were
# merged into a single ngen catchment), with `link`/`to` holding the raw
# per-flowline COMID chain instead of the ngen id/toid. Every consumer of
# these tables (t-route, ngen's channel-routing config) expects exactly one
# row per flowpath id -- duplicate ids crash t-route's DataFrame.reindex()
# with "cannot reindex on an axis with duplicate labels". The reference
# hydrofabric has already reconciled this (link == id, to == toid, one row
# per flowpath), so do the same here: sum the additive length and
# length-weight-average the per-unit-length hydraulic parameters across the
# constituent flowlines making up each flowpath.

first_non_na <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[1]
}

# collapse a raw, flowline-level flowpath-attributes(-ml) table down to one
# row per `id`, aggregating `hydraulic_cols` as a Length_m-weighted mean and
# `Length_m` itself as a sum.
aggregate_flowpath_attributes <- function(df, hydraulic_cols) {
  df |>
    st_drop_geometry() |>
    group_by(id) |>
    summarise(
      toid             = first_non_na(toid),
      total_length_m   = sum(Length_m, na.rm = TRUE),
      across(all_of(hydraulic_cols), ~ weighted.mean(.x, w = Length_m, na.rm = TRUE)),
      alt              = first_non_na(alt),
      gage             = first_non_na(gage),
      gage_nex_id      = first_non_na(gage_nex_id),
      WaterbodyID      = first_non_na(WaterbodyID),
      waterbody_nex_id = first_non_na(waterbody_nex_id),
      vpuid            = first_non_na(vpuid),
      .groups = "drop"
    ) |>
    rename(Length_m = total_length_m) |>
    mutate(
      link = id,
      to   = toid,
      # these columns are all-NA for VPUs with no gaged/lake-associated
      # flowpaths, which makes first_non_na() return a logical NA; without an
      # explicit cast that gets written as BOOLEAN instead of the target
      # schema's REAL/TEXT type.
      alt              = as.double(alt),
      gage             = as.character(gage),
      gage_nex_id      = as.character(gage_nex_id),
      WaterbodyID      = as.character(WaterbodyID),
      waterbody_nex_id = as.character(waterbody_nex_id)
    )
}

flowpath_attributes <- st_read(params$ngen_gpkg, layer = "flowpath-attributes") |>
  aggregate_flowpath_attributes(
    hydraulic_cols = c("Y", "n", "nCC", "BtmWdth", "TopWdth", "TopWdthCC",
                        "ChSlp", "So", "MusX", "MusK")
  )

flowpath_attributes_ml <- st_read(params$ngen_gpkg, layer = "flowpath-attributes-ml") |>
  aggregate_flowpath_attributes(
    hydraulic_cols = c("n", "BtmWdth", "TopWdth", "ChSlp", "nCC", "TopWdthCC",
                        "Y", "YCC", "dingman_r", "So", "MusX", "MusK")
  )

```

## Write the updated divides, flowpaths, nexus, network, pois, hydrolocations, lakes, and flowpath attributes to the output NextGen Geopackage

```{r}

# replace existing layers in the output NextGen Geopackage with the updated versions
st_write(divides,        params$out_nextgen, layer = "divides",        delete_layer = TRUE)
st_write(flowpaths,      params$out_nextgen, layer = "flowpaths",      delete_layer = TRUE)
st_write(nexus,          params$out_nextgen, layer = "nexus",          delete_layer = TRUE)
st_write(network,        params$out_nextgen, layer = "network",        delete_layer = TRUE)
st_write(pois,           params$out_nextgen, layer = "pois",           delete_layer = TRUE)
st_write(hydrolocations, params$out_nextgen, layer = "hydrolocations", delete_layer = TRUE)
st_write(lakes,          params$out_nextgen, layer = "lakes",          delete_layer = TRUE)
st_write(flowpath_attributes,    params$out_nextgen, layer = "flowpath-attributes",    delete_layer = TRUE)
st_write(flowpath_attributes_ml, params$out_nextgen, layer = "flowpath-attributes-ml", delete_layer = TRUE)

# drop intermediate layers that are byproducts of build_nextgen and are not
# part of the target NextGen data model
out_layers <- st_layers(params$out_nextgen)$name
for (lyr in intersect(c("flowlines", "incremental_areas"), out_layers)) {
  st_delete(params$out_nextgen, layer = lyr)
}

```

## Enforce Physical Column Order to Match ngiab_data_cli's Expected Schema

```{r}

# ngiab_data_cli's subset_table()/insert_data() copy rows between geopackages
# with `SELECT * FROM source` and a *positional* `INSERT INTO dest VALUES (...)`,
# with `dest` pre-created from a bundled template.sql that fixes each table's
# physical column order. sf::st_write() does not guarantee our column order
# matches that template (and always forces "geom" to sit right after "fid"),
# so any mismatch silently shifts values into the wrong columns downstream.
# Rebuild each feature table here with columns in the exact order
# ngiab_data_cli's template.sql expects, using name-based SELECT/INSERT so
# values always land in the correct destination column regardless of the
# table's current physical order.

library(DBI)
library(RSQLite)

# fid is always first; "geom" position is included explicitly below since it
# is not consistently placed by ngiab_data_cli's template (e.g. it's last in
# hydrolocations instead of second).
target_column_order <- list(
  "flowpaths" = c("fid", "geom", "id", "toid", "mainstem", "order", "hydroseq",
                   "lengthkm", "areasqkm", "tot_drainage_areasqkm", "has_divide",
                   "divide_id", "poi_id", "vpuid"),
  "divides" = c("fid", "geom", "divide_id", "toid", "type", "ds_id", "areasqkm",
                "vpuid", "id", "lengthkm", "tot_drainage_areasqkm", "has_flowline"),
  "lakes" = c("fid", "geom", "lake_id", "LkArea", "LkMxE", "WeirC", "WeirL",
              "OrificeC", "OrificeA", "OrificeE", "WeirE", "ifd", "Dam_Length",
              "domain", "poi_id", "hf_id", "reservoir_index_AnA",
              "reservoir_index_Extended_AnA", "reservoir_index_GDL_AK",
              "reservoir_index_Medium_Range", "reservoir_index_Short_Range",
              "res_id", "vpuid", "lake_x", "lake_y"),
  "pois" = c("fid", "poi_id", "id", "nex_id", "vpuid"),
  "hydrolocations" = c("fid", "poi_id", "id", "nex_id", "hf_id", "hl_link",
                        "hl_reference", "hl_uri", "hl_source", "hl_x", "hl_y",
                        "vpuid", "geom"),
  "flowpath-attributes" = c("fid", "link", "to", "Length_m", "Y", "n", "nCC",
                             "BtmWdth", "TopWdth", "TopWdthCC", "ChSlp", "alt",
                             "So", "MusX", "MusK", "gage", "gage_nex_id",
                             "WaterbodyID", "waterbody_nex_id", "id", "toid",
                             "vpuid"),
  "flowpath-attributes-ml" = c("fid", "link", "to", "Length_m", "alt", "So",
                                "MusX", "MusK", "gage", "gage_nex_id",
                                "WaterbodyID", "waterbody_nex_id", "id", "toid",
                                "vpuid", "n", "BtmWdth", "TopWdth", "ChSlp",
                                "nCC", "TopWdthCC", "Y", "YCC", "dingman_r"),
  "network" = c("fid", "id", "toid", "divide_id", "ds_id", "mainstem",
                "hydroseq", "hf_source", "hf_id", "lengthkm", "areasqkm",
                "tot_drainage_areasqkm", "type", "vpuid", "hf_hydroseq",
                "hf_lengthkm", "hf_mainstem", "topo", "poi_id", "hl_uri"),
  "nexus" = c("fid", "geom", "id", "toid", "type", "vpuid", "poi_id"),
  "divide-attributes" = c(
    "fid", "divide_id", "mode.bexp_soil_layers_stag=1", "mode.bexp_soil_layers_stag=2",
    "mode.bexp_soil_layers_stag=3", "mode.bexp_soil_layers_stag=4", "mode.ISLTYP",
    "mode.IVGTYP", "geom_mean.dksat_soil_layers_stag=1", "geom_mean.dksat_soil_layers_stag=2",
    "geom_mean.dksat_soil_layers_stag=3", "geom_mean.dksat_soil_layers_stag=4",
    "geom_mean.psisat_soil_layers_stag=1", "geom_mean.psisat_soil_layers_stag=2",
    "geom_mean.psisat_soil_layers_stag=3", "geom_mean.psisat_soil_layers_stag=4",
    "mean.cwpvt", "mean.mfsno", "mean.mp", "mean.refkdt", "mean.slope_1km",
    "mean.smcmax_soil_layers_stag=1", "mean.smcmax_soil_layers_stag=2",
    "mean.smcmax_soil_layers_stag=3", "mean.smcmax_soil_layers_stag=4",
    "mean.smcwlt_soil_layers_stag=1", "mean.smcwlt_soil_layers_stag=2",
    "mean.smcwlt_soil_layers_stag=3", "mean.smcwlt_soil_layers_stag=4",
    "mean.vcmx25", "mean.Coeff", "mean.Zmax", "mode.Expon", "centroid_x",
    "centroid_y", "mean.impervious", "mean.elevation", "mean.slope",
    "circ_mean.aspect", "dist_4.twi", "vpuid"
  )
)

# Some columns come through from the upstream build with the wrong SQL type
# (e.g. all-NA/ID-like columns that GDAL infers as REAL instead of the
# TEXT they should be) and are never otherwise rewritten by this script, so
# they can't be fixed via mutate()/st_write(). Force their declared type here
# regardless of what PRAGMA table_info() currently reports.
column_type_overrides <- list(
  "flowpath-attributes"    = c(link = "TEXT", to = "TEXT"),
  "flowpath-attributes-ml" = c(link = "TEXT", to = "TEXT")
)

con <- dbConnect(RSQLite::SQLite(), params$out_nextgen)

for (tbl in names(target_column_order)) {
  if (!(tbl %in% dbListTables(con))) next

  target_cols <- target_column_order[[tbl]]
  current_cols <- dbListFields(con, tbl)

  # only reorder if the column sets match exactly (same names, different order);
  # if they don't match, leave the table untouched so an unexpected schema
  # drift doesn't silently drop data.
  if (!setequal(target_cols, current_cols)) {
    warning(paste0(
      "Skipping column reorder for '", tbl, "': column sets differ.\n",
      "  expected: ", paste(sort(target_cols), collapse = ", "), "\n",
      "  actual:   ", paste(sort(current_cols), collapse = ", ")
    ))
    next
  }

  # get the current column types so the rebuilt table preserves them
  info <- dbGetQuery(con, sprintf('PRAGMA table_info("%s")', tbl))
  col_types <- setNames(info$type, info$name)

  overrides <- column_type_overrides[[tbl]]
  if (!is.null(overrides)) {
    col_types[names(overrides)] <- overrides
  }

  # skip the rebuild entirely only if both the column order already matches
  # the target AND no declared type needs to change
  types_changed <- !is.null(overrides) &&
    !identical(unname(col_types[names(overrides)]), unname(setNames(info$type, info$name)[names(overrides)]))
  if (identical(target_cols, current_cols) && !types_changed) next

  quote_id <- function(x) paste0('"', x, '"')
  col_defs <- vapply(target_cols, function(col) {
    if (col == "fid") {
      paste0(quote_id(col), " INTEGER NOT NULL")
    } else {
      paste0(quote_id(col), " ", col_types[[col]])
    }
  }, character(1))

  tmp_tbl <- paste0(tbl, "__reorder_tmp")
  dbExecute(con, sprintf('DROP TABLE IF EXISTS %s', quote_id(tmp_tbl)))
  dbExecute(con, sprintf(
    'CREATE TABLE %s (%s, PRIMARY KEY("fid" AUTOINCREMENT))',
    quote_id(tmp_tbl), paste(col_defs, collapse = ", ")
  ))

  col_list <- paste(vapply(target_cols, quote_id, character(1)), collapse = ", ")
  dbExecute(con, sprintf(
    'INSERT INTO %s (%s) SELECT %s FROM %s',
    quote_id(tmp_tbl), col_list, col_list, quote_id(tbl)
  ))

  dbExecute(con, sprintf('DROP TABLE %s', quote_id(tbl)))
  dbExecute(con, sprintf('ALTER TABLE %s RENAME TO %s', quote_id(tmp_tbl), quote_id(tbl)))
}

dbDisconnect(con)

```

## Register Non-Spatial Attribute Tables in gpkg_contents

```{r}

# sf::st_write() (and the upstream build step that originally produced
# divide-attributes/flowpath-attributes/flowpath-attributes-ml) do not
# register non-spatial ("aspatial") tables in gpkg_contents. GDAL/ogrinfo
# still lists them as layers via a fallback heuristic, but any tool that
# enumerates layers strictly via gpkg_contents (e.g. ngiab_data_cli's
# get_feature_tables()/get_available_tables()) will not see them at all.
# Register any of our attribute-only tables that are missing, matching the
# convention used by the official reference hydrofabric geopackages
# (data_type = "attributes", srs_id = 0, no bounding box).
attribute_only_tables <- c(
  "network", "pois", "divide-attributes", "flowpath-attributes",
  "flowpath-attributes-ml"
)

con <- dbConnect(RSQLite::SQLite(), params$out_nextgen)

registered <- dbGetQuery(con, "SELECT table_name FROM gpkg_contents")$table_name
now_iso <- strftime(as.POSIXct(Sys.time(), tz = "UTC"), "%Y-%m-%dT%H:%M:%OS3Z")

for (tbl in attribute_only_tables) {
  if (!(tbl %in% dbListTables(con))) next
  if (tbl %in% registered) next

  dbExecute(con, "
    INSERT INTO gpkg_contents
      (table_name, data_type, identifier, description, last_change, srs_id)
    VALUES (?, 'attributes', ?, '', ?, 0)
  ", params = list(tbl, tbl, now_iso))
}

dbDisconnect(con)

```


