---
title: "run-hfngen"
output: html_document
date: "2026-04-08"
---

### Install packages

```{r}

install.packages("remotes")
install.packages("devtools", dependencies = TRUE)
install.packages("sf", dependencies = TRUE)

#remotes::install_github("lynker-spatial/hfngen", dependencies = TRUE)
install.packages('../lynkerspatial/hfngen', repos=NULL, type="source")
```

```{r}

# Install Arrow with Brotli support.
# Brotli is used to compress the channel hydraulic attributes parquet
# so it's needed to decompress them too.
Sys.setenv(
  ARROW_WITH_BROTLI = "ON",
  ARROW_USE_PKG_CONFIG = "true"
)

install.packages("arrow", type = "source")

```

```{r}

arrow::codec_is_available("brotli")
```
-   **Hydrofabric inputs** must include `flowpaths`, `divides`, and (optionally) flowline-resolution layers (`flowlines`, `incremental_areas`, `hydrolocations`).

-   **Raster inputs** should share a compatible CRS with the divides layer. The package will reproject geometries on the fly when needed, but aligning rasters ahead of time avoids repeated transforms.

-   **Hydraulic attributes** need a `feature_id` column that matches `network$reference_id`, the attributes you wish to aggregate, and a length-based weighting column.

```{r}

library(sf)

#library(hfngen)
devtools::load_all("../lynkerspatial/hfngen")
```

Define paths to input files

```{r}

flowpath_gpkg <- "vpu_01/vpu01_aggregate_distribution.gpkg"
flowline_gpkg <- "vpu_01/refactored_reference_hydrofabric.gpkg"

#st_layers(flowpath_gpkg)
#st_layers(flowline_gpkg)

```

```{r}

out_nextgen <- "vpu_01/vpu01_ngen.gpkg"

build_nextgen(flowpath_gpkg,
              flowline_gpkg = flowline_gpkg,
              outfile = out_nextgen)
```

```{r}

# gpkg: Path to a GeoPackage containing `divides` and `network` layers.
# - The `divides` layer must contain `divide_id` and valid geometries.
# - The `network` layer must contain: `divide_id`, `incremental_area_id`, `areasqkm`, `incremental_areasqkm`, and `reference_id`.
# ISLTYP: Path to a raster of soil type (ISLTYP).
# IVGTYP: Path to a raster of vegetation type (IVGTYP).
# IMP: Path to a raster for impervious fraction used for
# GW DuckDB connection via `hfutils::tbl_http()` with groundwater attributes,
#'   expected to include columns `ComID` and `Zmax`.



soil      <- "vpu_01/ISLTYP.tif"
veg       <- "vpu_01/IVGTYP.tif"
imperv    <- "vpu_01/CONUS_imp_240.tif"

gw_basins <- arrow::open_dataset("vpu_01/nwm_gw_basins.parquet")

minimal_div_atts(out_nextgen,
                 ISLTYP = soil,
                 IVGTYP = veg,
                 IMP    = imperv,
                 GW     = gw_basins)

```


```{r}

# network A data frame/tibble containing at least:
# - `reference_id` (IDs that link to the attribute dataset),
# - `hl_class` and `hl_reference` (to extract gage references),
# - the grouping column given by `hf_id` (default: `"flowline_id"`),
# - and (optionally) other columns used downstream.
# This is typically the `network` layer produced by your NextGen build.
net = sf::read_sf(out_nextgen, "network")


# hydraulics Arrow dataset or DuckDB/DBI table (e.g., via
# `hfutils::tbl_http()`) containing:
# - `feature_id` (mapped to `reference_id`),
# - the columns listed in `attribute_cols`,
# - the weighting column named in `weight_col`.
hydraulic_atts <- arrow::open_dataset("vpu_01/riverml_channel_geometry_with_ahg.parquet")

```

```{r}
atts = c('owp_y_bf', 'owp_tw_bf','bf_area', 'owp_roughness_bathy','slope')
calculate_flow_atts(net,
                    hydraulic_atts,
                    hf_id = "flowline_id",
                    attribute_cols = atts,
                    weight_col = 'lengthkm')
```
